---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

```{r setoptions, echo=FALSE}
knitr::opts_chunk$set(echo=TRUE, message=FALSE)
```

## Loading and preprocessing the data

Load the raw data
```{r}
dat <- read.csv("activity.csv")
```

## What is mean total number of steps taken per day?

Firts I calculate the total number of steps taken per day and store these in a new variable. I omit NAs from the dataset.  
```{r}
library(dplyr)
steps_per_day <- filter(dat, !is.na(dat$steps)) # Remove NAs
steps_per_day <- group_by(steps_per_day, date) # Group by day
steps_per_day <- summarize(steps_per_day, steps=sum(steps)) # Calculate steps per day
```
  
Then I make a histrogram of the total number of steps taken each day.
```{r}
library(ggplot2)
ggplot(steps_per_day, aes(x=steps)) + geom_histogram()
```
  
Finally I calculate the mean and median total number of steps taken per day.
```{r}
mean <- mean(steps_per_day$steps)
median <- median(steps_per_day$steps)
```
mean: `r format(mean, digits = 1)`  
median: `r format(median, digits=1)`

## What is the average daily activity pattern?

Firts I calculate the mean steps taken per interval and store these in a new variable. I omit NAs from the dataset.  
```{r}
library(dplyr)
steps_per_interval <- filter(dat, !is.na(dat$steps)) # Remove NAs
steps_per_interval <- group_by(steps_per_interval, interval) # Group by interval
steps_per_interval <- summarize(steps_per_interval, steps=mean(steps)) # Calculate steps mean per interval
```

Then I plot the results in a line-graph.
```{r}
ggplot(steps_per_interval, aes(x=interval, y=steps)) + geom_line()
```

Finally, I calculate which interval has the highest steps average.
```{r}
library(dplyr)
max_steps <- max(steps_per_interval$steps)
max_interval <- select(filter(steps_per_interval, steps==max_steps), interval)
```
Maximum average steps are `r format(max_steps, digits=1)`, which correspond to the interval `r format(max_interval, digits=1)`.

## Imputing missing values
The total number of missing values in the dataset
```{r}
sum(is.na(dat$steps))
```

Create a function that assigns the mean value of that interval whenever an NA is present.
```{r}
replace_NA <- function(steps_per_interval, NA_steps, NA_interval){
    if(is.na(NA_steps)){
        as.numeric(select(filter(steps_per_interval, interval==NA_interval), steps))
    } else NA_steps 
}
```

Create a new dataframe with replacement for NA values.
```{r}
dat_complete <- dat
dat_complete$steps <- sapply(1:nrow(dat_complete), function(x) replace_NA(steps_per_interval, dat$steps[x], dat$interval[x]))
```

Using the adjusted dataframe, I calculate the total number of steps taken per day and store these in a new variable.
```{r}
library(dplyr)
steps_per_day <- group_by(dat_complete, date) # Group by day
steps_per_day <- summarize(steps_per_day, steps=sum(steps)) # Calculate steps per day
```
  
Using the adjusted dataframe, I make a histrogram of the total number of steps taken each day.
```{r}
library(ggplot2)
ggplot(steps_per_day, aes(x=steps)) + geom_histogram()
```
  
Using the adjusted dataframe, I calculate the mean and median total number of steps taken per day.
```{r}
mean_complete <- mean(steps_per_day$steps)
median_complete <- median(steps_per_day$steps)
```
mean: `r format(mean_complete, digits = 1)`  
median: `r format(median_complete, digits=1)`
  
  
Replacing the NA values does not significantly change the mean and median:
Mean: `r format(mean_complete, digits = 7)` vs. `r format(mean, digits = 7)`     
Median: `r format(median_complete, digits=7)` vs. `r format(median, digits=7)`

## Are there differences in activity patterns between weekdays and weekends?

Create new factor variable with levels "weekday" and "weekend".
```{r}
dat_complete$weekday <- as.factor(sapply(1:nrow(dat_complete), function(x) if(as.POSIXlt(dat$date[x])$wday>5) "weekend" else "weekday"))
```

Make panel plot of weekdays and weekends.
```{r}
ggplot(dat_complete, aes(x=interval, y=steps)) + geom_line(stat="summary", fun.y="mean") + facet_wrap(~weekday, ncol=1)
```

